---
title: "Final Project"
author: "Gabe Riedel, Oliver Whitmarsh"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
code-fold: true
---


Propose a Bayesian model - Oliver
Discuss choice of likelihood and assumptions of your model - Oliver
Discuss choice of prior, including some prior predictive tuning - Oliver
Fit the model to the data - Oliver
Perform and summarize posterior inference - Gabe
Perform and summarize posterior prediction - Gabe
Perform and summarize posterior predictive checking or model comparison - Gabe
Perform sensitivity analysis (e.g., how does the posterior change if the prior changes) - Gabe
Compare your results with a comparable frequentist analysis - Oliver
Report conclusions in the context of the research question - Gabe


```{r}
#| output: false
library(tidyverse)
library(brms)
```


## Research Question and Context

Tunneling is a concept in baseball theory that has been around for decades. It is the idea that two pitches can look very similar to eye of batter, until they don't. Pitching is all about deceiving the batter, and if a pitcher has pitches in their arsenal that "tunnel" well off of each other, then they are more likely to have success getting hitters out. So let's quantify this. Our question is which two-pitch sequences are the most effective in getting hitter's to swing and miss (i.e. which pitches tunnel the best off of each other)?  

## Data

```{r read-data}
cp_data <- read_csv("data/cp_bip.csv")
```

```{r clean-data}
cp_clean <- cp_data |>
  drop_na()
```

Our data come from the Cal Poly baseball team. They are every pitch thrown in the 2025 season by Cal Poly baseball pitchers. These data are recorded with a high speed radar system called Trackman that is present in majority of the college baseball stadiums in the country. Trackman produces CSVs at the end of each game where each row represents a pitch thrown and contains information on the ball flight of the pitch as well as the result of the play. Trackman is a highly accurate radar system designed for high speed ball tracking. Our data are a concatenation of the CSVs from each game in Cal Poly's 2025 season.


Predict Hit from ExitSpeed, Angle, and Direction

